\textbf{Unshuffling Permutations}

### 2016-04-07 slides/slides.tex
Suite

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame} \frametitle{Recognizing square words}
\begin{block}{Definition}
    A \alert{perfect matching} of a word $u$ is a graph $(V, E)$ such
    that
    \begin{itemize}
        \item $V := \{(u_i, i) : i \in \{1, \dots, |u|\}$;
        \item $(u_i, i) \-- (u_j, j) \in E$ implies $u_i = u_j$;
        \item every vertex of $V$ belongs to exactly one edge of $E$.
    \end{itemize}
\end{block}

@@

\begin{block}{Main questions}
    \begin{itemize}
        \item How the previous key results about the shuffle of words
        translate for the supershuffle of permutations?
        \medskip

        \item
    \end{itemize}
\end{block}
\end{frame}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Algebraic issues}

@@

\begin{frame}
  \frametitle{Permutation pattern}

  \begin{definition}
    A permutation $\pi$ is said to \emph{contain} the permutation $\sigma$ if there
    exists a subsequence of (not necessarily consecutive) entries of $\pi$ that
    has the same relative order as $\sigma$, and in this case $\sigma$ is said
    to be a pattern of $\pi$, written $\sigma \leq \pi$.

@@

    Otherwise, $\pi$ is said to \emph{avoid} the permutation $\sigma$.
  \end{definition}

@@

  \begin{exampleblock}{Example}
    \begin{itemize}
      \item
        The permutation $\pi = 391867452$ contains the pattern $\sigma = 51342$,
        as can be seen in the highlighted subsequence of
        $\pi = 3\mathbf{91}8\mathbf{674}52$
        (or $\pi = 3\mathbf{91}8\mathbf{67}4\mathbf{5}2$ or
        $\pi = 3\mathbf{91}8\mathbf{67}45\mathbf{2}$ or
        $\pi = 3\mathbf{91}867\mathbf{452}$.)

      \item
       Since the permutation $\pi = 391867452$ contains no increasing subsequence
       of length four, $\pi$ avoids $1234$.
    \end{itemize}
  \end{exampleblock}

@@

\begin{frame}
  \frametitle{Square permutation}

  \begin{definition}
    A permutation is said to be a \emph{square} if it can be obtained by
    shuffling two order-isomorphic patterns
  \end{definition}

@@

  \begin{exampleblock}{Example}
    \begin{itemize}
      \item
        The permutation $\pi = 25167834$ is a square since it can obtained by
        shuffling two order-isomorphic (\emph{i.e.} $1342$)
        patterns $1683$ and $2574$ as can be seen in the highlighted subsequence of
        $\pi = \mathbf{25}16\mathbf{7}83\mathbf{4}$.

@@

      \item
       $\pi = 4123$ and $1432$ are not squares.
    \end{itemize}
  \end{exampleblock}

@@

\begin{frame}
  \frametitle{Hardness}
  \framesubtitle{Oriented perfect matching}

  \begin{definition}[Property $\mathbf{P_1}$]
    Let $\pi$ be a permutation. An oriented perfect matching $\mathcal{M}$
    on $\pi$ is said to have property $\mathbf{P_1}$ if it avoids all the
    six patterns \InclusionLL, \InclusionLR, \InclusionRL,
    \InclusionRR, \CrossingLR, and \CrossingRL.
  \end{definition}

  \begin{definition}[Property $\mathbf{P_2}$]
    Let $\pi$ be a permutation. An oriented perfect matching $\mathcal{M}$
    on $\pi$ is said to have property $\mathbf{P_2}$ if, for any two
    distinct arcs $(\pi(a), \pi(a'))$ and $(\pi(b), \pi(b'))$ in $\mathcal{M}$,
    we have $\pi(a) < \pi(b)$ if and only if $\pi(a') < \pi(b')$.
  \end{definition}
\end{frame}

\begin{frame}
  \frametitle{Hardness}
  \framesubtitle{Oriented perfect matching}

  \begin{theorem}
    Let $\pi$ be a permutation. The following statements are equivalent:
    \begin{enumerate}
      \item The permutation $\pi$ is a square.
      \item There exists an oriented perfect matching $\mathcal{M}$
      on $\pi$ satisfying~$\mathbf{P_1}$ and~$\mathbf{P_2}$.
    \end{enumerate}
  \end{theorem}

  \onslide<2->
  \begin{exampleblock}{Example}
    \begin{tikzpicture}[scale=0.35,inner sep=2pt,node distance=1cm]
        \draw node [draw,circle,minimum size=.55cm,fill=black!10](U01){$1$};
        \draw node [draw,circle,minimum size=.55cm,fill=black!10,right of=U01]
            (U02) {$8$};
        \draw node [draw,circle,minimum size=.55cm,right of=U02](U03){$3$};
        \draw node [draw,circle,minimum size=.55cm,fill=black!10,right of=U03]
            (U04) {$9$};
        \draw node [draw,circle,minimum size=.55cm,fill=black!10,right of=U04]
            (U05) {$2$};
        \draw node [draw,circle,minimum size=.55cm,right of=U05](U06){$7$};
        \draw node [draw,circle,minimum size=.55cm,right of=U06](U07){$B$};
        \draw node [draw,circle,minimum size=.55cm,right of=U07](U08){$5$};
        \draw node [draw,circle,minimum size=.55cm,right of=U08](U09){$C$};
        \draw node [draw,circle,minimum size=.55cm,right of=U09](U10){$6$};
        \draw node [draw,circle,minimum size=.55cm,fill=black!10,right of=U10]
            (U11) {$A$};
        \draw node [draw,circle,minimum size=.55cm,fill=black!10,right of=U11]
            (U12) {$4$};
        \draw [thick,->,>=latex']
        (U01.north) .. controls ($ (U01.north) + (0,2.25) $)
            and ($ (U03.north) + (0,2.25) $) .. (U03.north);
        \draw [thick,->,>=latex']
        (U02.north) .. controls ($ (U02.north) + (0,2.25) $)
            and ($ (U06.north) + (0,2.25) $) .. (U06.north);
        \draw [thick,->,>=latex']
        (U04.north) .. controls ($ (U04.north) + (0,2.5) $)
            and ($ (U07.north) + (0,2.5) $) .. (U07.north);
        \draw [thick,->,>=latex']
        (U05.north) .. controls ($ (U05.north) + (0,2.5) $)
            and ($ (U08.north) + (0,2.5) $) .. (U08.north);
        \draw [thick,->,>=latex']
        (U11.north) .. controls ($ (U11.north) + (0,2.5) $)
            and ($ (U09.north) + (0,2.5) $) .. (U09.north);
        \draw [thick,->,>=latex']
        (U12.north) .. controls ($ (U12.north) + (0,2.5) $)
            and ($ (U10.north) + (0,2.5) $) .. (U10.north);
    \end{tikzpicture}

    {\small%
    An oriented perfect matching $\mathcal{M}$ on the permutation
    $\pi = 183927B5C6A4$ satisfying the properties $\mathbf{P_1}$
    and $\mathbf{P_2}$. From $\mathcal{M}$, it follows that $\pi$ is
    a square as it appears in the shuffle of $1892A4$ and $37B5C6$,
    both being order-isomorphic to $145263$.
    }%end small
  \end{exampleblock}
\end{frame}

\begin{frame}
  \frametitle{Hardness}

  \begin{theorem}
    Deciding whether a permutation is a square is \NPC.
  \end{theorem}

  \begin{block}{Remarks}
    \begin{itemize}
      \item
      We reduce from the problem of deciding wether a pattern $\sigma$ occurs in a
      permutation $\pi$.

      \smallskip

      \item
      Construct a new permutation $\mu$ such that
      $\sigma$ occurs in $\pi$ if and only if
      there exists an oriented perfect matching $\mathcal{M}$ on $\mu$
      satisfying $\mathbf{P_1}$ and $\mathbf{P_2}$.
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Hardness}

@@

  \begin{block}{Combinatorics}
    How many permutations of length $2n$ are squares?

    \smallskip

    How many $(213,231)$-avoiding
    permutations of length $2n$ are squares? (Equivalently,
    how many binary strings of length $2n$ are squares?)
  \end{block}

  \begin{block}{Combinatorics}
    Given two permutations $\pi$ and~$\sigma$, how hard is the problem of
    deciding whether $\sigma$ is a square root of~$\pi$?
  \end{block}

  \begin{block}{Algebra}
    \ldots
  \end{block}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame} \frametitle{Recognizing square words}
\begin{block}{Definition}
    A \alert{perfect matching} of a word $u$ is a graph $(V, E)$ such
    that
    \begin{itemize}
        \item $V := \{(u_i, i) : i \in \{1, \dots, |u|\}$;
        \item every vertex of $V$ belongs to exactly one edge of $E$;
        \item $(u_i, i) \-- (u_j, j) \in E$ implies $u_i = u_j$.
    \end{itemize}
\end{block}

@@

\begin{block}{Main questions}
    \begin{itemize}
        \item Combinatorial properties of ``square permutations''
        w.r.t. the supershuffle?
        \medskip

        \item Complexity of the recognition of ``square permutations''?
    \end{itemize}
\end{block}
\end{frame}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Algebraic and combinatorial issues}

@@

\begin{frame}
\frametitle{Permutation patterns}

  \begin{definition}
    A permutation $\pi$ \alert{contains} a permutation $\sigma$ if there
    exists a subsequence of (not necessarily consecutive) letters of $\pi$ that
    has the same relative order as $\sigma$, and in this case $\sigma$ is said
    to be a \alert{pattern} of $\pi$, written $\sigma \leq \pi$.
    \smallskip

    Otherwise, $\pi$ \alert{avoids} $\sigma$.
  \end{definition}

@@

  \begin{exampleblock}{Example}

        The permutation
        $\pi = 3\textcolor{BrickRed}{91}8\textcolor{BrickRed}{674}52$
        contains the pattern $\sigma := 51342$ since
        $\STD(\textcolor{BrickRed}{91674}) = 51342$.
  \end{exampleblock}
  \medskip

  \begin{block}{Definition}
    The \alert{pattern involvement problem} is the decision problem
    consisting in, given $\pi, \sigma \in S$, decide whether $\sigma \leq \pi$.
  \end{block}
  \medskip

  \begin{block}{Proposition \BIB{Bose, Buss, Lubiw}{1998}}
    The pattern involvement problem is $\NPC$.
  \end{block}
\end{frame}

%\begin{frame}
  %\frametitle{Square permutation}

  %\begin{definition}
    %A permutation is said to be a \emph{square} if it can be obtained by
    %shuffling two order-isomorphic patterns
  %\end{definition}

  %\medskip

  %\begin{exampleblock}{Example}
    %\begin{itemize}
      %\item
        %The permutation $\pi = 25167834$ is a square since it can obtained by
        %shuffling two order-isomorphic (\emph{i.e.} $1342$)
        %patterns $1683$ and $2574$ as can be seen in the highlighted subsequence of
        %$\pi = \mathbf{25}16\mathbf{7}83\mathbf{4}$.

      %\medskip

      %\item
       %$\pi = 4123$ and $1432$ are not squares.
    %\end{itemize}
  %\end{exampleblock}

%\end{frame}

\begin{frame}
  \frametitle{Recognizing square permutations}

    \begin{block}{Definition}
        A \alert{perfect matching} of a permutation $\pi \in S_n$ is a
        directed graph $(V, A)$ such that
        \begin{itemize}
            \item $V := \{(\pi_i, i) : i \in \{1, \dots, n\}\}$;
            \item every vertex of $V$ belongs to exactly one arc of $A$.
        \end{itemize}
    \end{block}
    \medskip

  \begin{definition}[Property $\mathbf{P_1}$]
    Let $\pi \in S$. A perfect matching $(V, A)$ of $\pi$ has property
    $\mathbf{P_1}$ if it avoids all the six patterns
    \InclusionLL, \InclusionLR, \InclusionRL,
    \InclusionRR, \CrossingLR, and \CrossingRL.
  \end{definition}

@@

  \begin{definition}[Property $\mathbf{P_2}$]
    Let $\pi \in S$. A perfect matching $(V, A)$ of $\pi$ has property
    $\mathbf{P_2}$ if, for all arcs $(\pi_i, i) \to (\pi_j, j)$ and
    $(\pi_k, k) \to (\pi_\ell, \ell)$ of $A$, $\pi_i < \pi_k$ iff
    $\pi_j < \pi_\ell$.
  \end{definition}
\end{frame}

@@

\begin{frame}
  \frametitle{Recognizing square permutations}

  \begin{block}{Proposition}
    A permutation $\pi$ is a square (w.r.t. the shuffle product $\SHUFFLE$)
    iff $\pi$ admits a perfect matching satisfying~$\mathbf{P_1}$
    and~$\mathbf{P_2}$.
  \end{block}

@@

  \onslide<2->
  \begin{exampleblock}{Example}
    The permutation $\pi := 183927B5C6A4$ is a square since it admits the
    perfect matching
    \begin{equation*}
    \begin{split}
    \begin{tikzpicture}[scale=0.35,inner sep=2pt,node distance=1cm]
        \tikzstyle{Vertex}=[draw,circle,minimum size=.45cm,font=\footnotesize]
        \draw node[Vertex,fill=RoyalBlue!30](U01){$1$};
        \draw node[Vertex,fill=RoyalBlue!30,right of=U01](U02){$8$};
        \draw node[Vertex,fill=BrickRed!30,right of=U02](U03){$3$};
        \draw node[Vertex,fill=RoyalBlue!30,right of=U03](U04){$9$};
        \draw node[Vertex,fill=RoyalBlue!30,right of=U04](U05){$2$};
        \draw node [Vertex,fill=BrickRed!30,right of=U05](U06){$7$};
        \draw node [Vertex,fill=BrickRed!30,right of=U06](U07){$B$};
        \draw node [Vertex,fill=BrickRed!30,right of=U07](U08){$5$};
        \draw node [Vertex,fill=BrickRed!30,right of=U08](U09){$C$};
        \draw node [Vertex,fill=BrickRed!30,right of=U09](U10){$6$};
        \draw node [Vertex,fill=RoyalBlue!30,right of=U10](U11){$A$};
        \draw node [Vertex,fill=RoyalBlue!30,right of=U11](U12){$4$};
        \draw [thick,->,>=latex']
        (U01.north) .. controls ($ (U01.north) + (0,2.25) $)
            and ($ (U03.north) + (0,2.25) $) .. (U03.north);
        \draw [thick,->,>=latex']
        (U02.south) .. controls ($ (U02.south) + (0,-2.25) $)
            and ($ (U06.south) + (0,-2.25) $) .. (U06.south);
        \draw [thick,->,>=latex']
        (U04.north) .. controls ($ (U04.north) + (0,2.5) $)
            and ($ (U07.north) + (0,2.5) $) .. (U07.north);
        \draw [thick,->,>=latex']
        (U05.south) .. controls ($ (U05.south) + (0,-2.5) $)
            and ($ (U08.south) + (0,-2.5) $) .. (U08.south);
        \draw [thick,->,>=latex']
        (U11.north) .. controls ($ (U11.north) + (0,2.5) $)
            and ($ (U09.north) + (0,2.5) $) .. (U09.north);
        \draw [thick,->,>=latex']
        (U12.north) .. controls ($ (U12.north) + (0,2.5) $)
            and ($ (U10.north) + (0,2.5) $) .. (U10.north);
    \end{tikzpicture}
    \end{split}
    \end{equation*}
    satisfying $\mathbf{P_1}$ and~$\mathbf{P_2}$.
    \smallskip

    Moreover, this shows that $\pi$ appears in the shuffle of the
    words $\textcolor{RoyalBlue}{1892A4}$ and
    $\textcolor{BrickRed}{37B5C6}$, both order isomorphic to
    $145263$.
  \end{exampleblock}
\end{frame}

\begin{frame}
  \frametitle{Hardness of recognizing square permutations}

  \begin{block}{Proposition}
    Deciding whether a permutation is a square is \NPC.
  \end{block}

@@

  \begin{block}{Main ideas of the proof}
    \begin{itemize}
        \item We show that the pattern involvement problem is reducible
        in polynomial time to our decision problem.
        \medskip

        \item For this, given two permutations $\sigma$ and $\pi$, we
        show that $\sigma \leq \pi$ iff $\mu_{\sigma, \pi}$ is a square,
        where $\mu_{\sigma, \pi}$ is a permutation constructed in polynomial
        time from $\sigma$ and $\pi$.
        \medskip

        \item To prove that $\mu_{\sigma, \pi}$ is a square, we show
        that $\mu_{\sigma, \pi}$ admits a perfect matching satisfying
        $\mathbf{P_1}$ and~$\mathbf{P_2}$.

    \end{itemize}
  \end{block}
\end{frame}

@@

\begin{frame}
  \frametitle{Hardness of recognizing square permutations}

@@

  \begin{block}{Combinatorics}
    \begin{itemize}
        \item Enumeration of square permutations?
        \smallskip

        \item Enumeration of square permutations avoiding $213$ and $231$?
    \end{itemize}
  \end{block}
  \medskip

  \begin{block}{Algorithms}

    Given two permutations $\pi$ and~$\sigma$, how hard is the problem of
    deciding whether $\sigma$ is a square root of~$\pi$?
  \end{block}
  \medskip

  \begin{block}{Algebra}
        How algebraic properties of the associative algebra $(\QQ[S], \SHUFFLE)$
        or of the coalgebra $(\QQ[S], \Delta)$ are connected to
        combinatorial and algorithmic properties of square permutations?
  \end{block}
